---
title: "tableHTML"
author: "Theo Boutaris"
date: "`r Sys.Date()`"
output: 
  html_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a package for building CSS-ible HTML tables in an easy way, which are compatible with R shiny. 
The main function `tableHTML` will convert a data.frame or matrix or any other object that can be
converted into a data.frame into an HTML table. Using the function on R Studio will print the table
on the viewer. The default tables are built without any CSS in order to allow for full flexibility
although later developements will include themes.

In order to see the most basic example with tableHTML you could type:

```{r basic}
library(tableHTML)
tableHTML(mtcars)
```

### Removing Row Names

To remove row names from the HTML table you can type:

```{r rownames}
tableHTML(mtcars, rownames = FALSE)
```

### Adding a table class

The class argument sets the class name for the table. The default will be of the form 
table\_dataframeName. For example, for mtcars the HTML table's class would be table\_mtcars.
This makes it easy to create a css value for it in case there are multiple tables.

```{r, eval = FALSE}
mytable <- tableHTML(mtcars)
str(mytable)
Classes 'tableHTML', 'html', 'character'  atomic [1:1] 
<table class=table_mtcars border=1 style="border-collapse: collapse;">
<tr>
  <th id=header_1> </th>
  <th id=header_2>mpg</th>
  <th id=header_3>cyl</th>
truncated...
```

And with a custom class:

```{r, eval = FALSE}
mytable <- tableHTML(mtcars, class = 'myClass')
str(mytable)
Classes 'tableHTML', 'html', 'character'  atomic [1:1] 
<table class=myClass border=1 style="border-collapse: collapse;">
<tr>
  <th id=header_1> </th>
  <th id=header_2>mpg</th>
  <th id=header_3>cyl</th>
truncated...
```


### Adding Second Headers that span multiple columns

In case you would like to add second headers that span multiple columns you can do it by providing a
list in the second_header argument. The first element will contain the column spans 
(i.e. a numeric atomic vector) whereas the second element will contain the names 
(i.e. a character atomic vector).

For example:

```{r second header}
library(tableHTML)
tableHTML(mtcars, second_header = list(c(3, 4, 5), c('col1', 'col2', 'col3')))
```

### Change the widths

Obviously, the above tables are far from appealing since they have no css. So we can start styling
them now. A very important feature is to be able to control the column widths. You can do this with
the widths argument. 

The widths argument specifies the columns' width in pixels. It needs to have the same length as the
number of columns (if rownames is TRUE, rownames are also considered a column). 
If second headers exist these will stretch according to their underlying columns' widths. 

As an example:

```{r widths}
tableHTML(mtcars, widths = rep(100, 12), second_header = list(c(3, 4, 5), c('col1', 'col2', 'col3')))
```


## add\_css\_* function family

These functions add inline CSS to the HTML table. The package has been developed so that the 
functions are chained with the `%>%` (pipe) operator which comes from the `magrittr` package. 
You can chain as many of those as you want
when building your HTML table. The CSS code is added within the HTML table. This is ok for simple
table styling but for more complex CSS styles (like the hover selector) it would still require a
separate css file (we will see how to work with those in shiny later on).

### add\_css\_row

In order to add css in specific table rows you need to use the `add_css_row` function. The function
has three arguments:

1. *tableHTML*: A tableHTML object (created by `tableHTML()`)
2. *css*: A list of two elements with the corresponding css. The first element of the list should 
          be an atomic vector with the style definitions (e.g. background-color). The second element
          will be an atomic vector with the same length as the first element, which will contain the
          style definitions' values (e.g. red).
3. *rows*: A numeric atomic vector with the indices of the rows on which the style definitions 
           will be applied. **headers and second_headers are also part of the rows**. Default is NULL
           which means that it will be applied to all rows.

For example:

```{r add css row 1}
mtcars %>%
  tableHTML(widths = c(140, rep(45, 11))) %>% 
  add_css_row(css = list(c('background-color', 'border'), c('lightgray', '3px solid green')))
```

\    

And adding row indices manually:

```{r add css row 2}
mtcars %>%
  tableHTML(widths = c(140, rep(45, 11))) %>% 
  add_css_row(css = list(c('background-color', 'border'), c('lightgray', '3px solid green')),
              rows = 2:33)
```

\   

It is very common in HTML tables to highlight odd rows with one color and even rows with another.
We can achieve this with the rows argument and the helper funtions from the tableHTML package `odd`
and `even`.

```{r add css row 3}
mtcars %>%
  tableHTML(widths = c(140, rep(45, 11))) %>% 
  add_css_row(css = list('background-color', '#f2f2f2'),
              rows = odd(1:33)) %>%
  add_css_row(css = list('background-color', '#e6f0ff'),
              rows = even(1:33))
```

### add\_css\_column

In order to add css in specific table column you need to use the `add_css_column` function. 
The function has three arguments:

1. *tableHTML*: A tableHTML object (created by `tableHTML()`)
2. *css*: A list of two elements with the corresponding css. The first element of the list should 
          be an atomic vector with the style definitions (e.g. background-color). The second element
          will be an atomic vector with the same length as the first element, which will contain the
          style definitions' values (e.g. red).
3. *column_names*: A character atomic vector with the names of the columns where the style
                   definitions will be applied on. At least one column name must be provided. 
                   If the rownames are included the column name is "rownames". headers and second_headers
                   are not part of the columns. These need to be set from their corresponding add_css 
                   functions.

For example:

```{r add css column 1}
mtcars %>%
  tableHTML(widths = c(140, rep(45, 11))) %>% 
  add_css_column(css = list(c('background-color', 'border'), c('lightgray', '3px solid green')), 
                 column_names = c('cyl', 'hp', 'rownames'))
```

\    

Keep in mind that if you apply the **same** style to both rows and columns, column styles will 
overwrite row styles:

```{r add css column 2}
mtcars %>%
  tableHTML(widths = c(140, rep(45, 11))) %>% 
  add_css_row(css = list('background-color', '#f2f2f2')) %>%
  add_css_column(css = list('background-color', 'lightblue'), 
                 column_names = c('cyl', 'hp', 'rownames')) 
 
```

### add\_css\_header

In order to add css in specific table header you need to use the `add_css_header` function. 
The function has three arguments:

1. *tableHTML*: A tableHTML object (created by `tableHTML()`)
2. *css*: A list of two elements with the corresponding css. The first element of the list should 
          be an atomic vector with the style definitions (e.g. background-color). The second element
          will be an atomic vector with the same length as the first element, which will contain the
          style definitions' values (e.g. red).
3. *headers*: A numeric atomic vector with the indices of the headers where the style definitions 
              will be applied on. At least one header index must be provided.

For example:

```{r add css header 1}
mtcars %>%
  tableHTML(widths = c(140, rep(45, 11))) %>% 
  add_css_header(css = list('background-color', 'lightgray'), headers = c(1, 4))
```

### add\_css\_second\_header

In order to add css in specific table second header you need to use the 
`add_css_second_header` function. The function has three arguments:

1. *tableHTML*: A tableHTML object (created by `tableHTML()`)
2. *css*: A list of two elements with the corresponding css. The first element of the list should 
          be an atomic vector with the style definitions (e.g. background-color). The second element
          will be an atomic vector with the same length as the first element, which will contain the
          style definitions' values (e.g. red).
3. *second_headers*: A numeric atomic vector with the indices of the second headers where the 
                     style definitions will be applied on. At least one second header index must 
                     be provided.
For example:

```{r add css second header 1}
mtcars %>%
  tableHTML(widths = c(140, rep(45, 11)), 
            second_header = list(c(3, 4, 5), c('col1', 'col2', 'col3'))) %>% 
  add_css_second_header(css = list(c('background-color', 'border'), 
                                   c('lightgray', '3px solid green')),
                         second_headers = c(1, 3))  
```

### All together

An example to put everything in action:

```{r all together}
mtcars %>%
  tableHTML(widths = c(140, rep(45, 11)),
            second_header = list(c(3, 4, 5), c('team1', 'team2', 'team3'))) %>% 
  add_css_second_header(css = list(c('height', 'background-color', 'font-size'), 
                                   c('40px', ' #e6e6e6', '30px')),
                        second_headers = 1:3) %>%
  add_css_header(css = list(c('height', 'background-color'), c('30px', ' #e6e6e6')),
                 headers = 1:12) %>%
  add_css_row(css = list('background-color', '#f2f2f2'),
              rows = even(1:34)) %>%
  add_css_row(css = list('background-color', '#e6f0ff'),
              rows = odd(1:34)) %>%
  add_css_column(css = list('text-align', 'center'), 
                 column_names = names(mtcars)) 
 
```


### Hover and other CSS selectors

These are not supported in HTML and they can be added through CSS files. We can see how to do this
in shiny below.

## tableHTML and shiny

Now that we know about how to use the function's arguments let's see how we can use tableHTML with 
shiny. The implementation is really simple since internally tableHTML uses `htmltools::HTML` to 
convert the table into an HTML object. Simply using `renderUI` and `uiOutput` will be enough:

```{r shiny 1, eval = FALSE}
shinyApp(
 ui = fluidPage(
   fluidRow(
    #leave some spacing
    br(),
    column(width = 1),
    uiOutput("mytable"))
 ), 
 server = function(input, output) {
  output$mytable <- renderUI( 
   tableHTML(mtcars)
 )}
)
```

Or with a more complex example:

```{r shiny 2, eval = FALSE}
shinyApp(
 ui = fluidPage(
   fluidRow(
    #leave some spacing
    br(),
    column(width = 1),
    uiOutput("mytable"))
 ), 
 server = function(input, output) {
  output$mytable <- renderUI( 
   mtcars %>%
     tableHTML(widths = c(140, rep(45, 11)),
               second_header = list(c(3, 4, 5), c('team1', 'team2', 'team3'))) %>% 
     add_css_second_header(css = list(c('height', 'background-color', 'font-size', 'text-align'), 
                                      c('40px', ' #e6e6e6', '30px', 'center')),
                           second_headers = 1:3) %>%
     add_css_header(css = list(c('height', 'background-color', 'text-align'), 
                               c('30px', ' #e6e6e6', 'center')),
                    headers = 1:12) %>%
     add_css_row(css = list('background-color', '#f2f2f2'),
                 rows = even(1:34)) %>%
     add_css_row(css = list('background-color', '#e6f0ff'),
                 rows = odd(1:34)) %>%
     add_css_column(css = list('text-align', 'center'), 
                    column_names = names(mtcars)) 
 )}
)
```


## shiny and CSS

As we said previously CSS selectors cannot be used within HTML so if you use shiny you would need
to include a separate CSS file. Before we start with shiny and CSS let's see the rules applied by
tableHTML. tableHTML creates ids for columns, headers and overheaders and gives the table a specific 
class name as it was mentioned previously.

The rules are like this:

* Table: Will get the class from the class argument in the function. The final class will be of 
  the form table_dataframeName.
* Columns: Will get an id which will be the same as the column name.
* Headers: Will get an id of the form header\_headerIndex. For example the first header will have 
  the id header\_1, the second header will have header\_2 and so on.
* Second\_Headers: Will get an id of the form overheader\_secondHeaderIndex. For example the first 
                   second\_header will have the id overheader\_1,
                   the second header will have overheader\_2 and so on.

Let's see this in practice. We create a CSS file e.g. mycss.css where we will write our css code. 
Then we load this in R shiny in the following way. We have saved a ui.R file and server.R file in 
the same directory. In that directory we have a www/ folder where we save our .css file. 
In our case that is named mycss.css. We include this file in R using `includeCSS`:

```{r shiny css, eval = FALSE}
#ui.R
shinyUI(
 fluidPage(
  fluidRow(
   #leave some spacing
   br(),
   column(width = 1),
   #include css file in shiny
   includeCSS('www/mycss.css'),
   uiOutput("mytable"))
 )
)

#server.R
shinyServer(
function(input, output) {
   output$mytable <- renderUI( 
    tableHTML(mtcars, second_header = list(c(3, 4, 5), c('col1', 'col2', 'col3')))
   )}
)
```

Note: You can find out more about CSS and shiny [here](http://shiny.rstudio.com/articles/css.html).

In order to add a vertical column line we can use the column ids created by tableHTML. To add a vertical red line at mpg column, our css file would look like this:

```{CSS}
#mpg {
	border-left: 5px solid red;
}
``` 


In order to change the background color of one of the headers (e.g. the second) our css would look like this:

```{CSS 2}
#mpg{
	border-left: 5px solid red;
}

#header_2 {
	background-color: green;
}
``` 


And in order to align the first of the second_headers in the center our css would look like:

```{CSS 3}
#mpg{
	border-left: 5px solid red;
}

#header_2 {
	background-color: green;
}

#overheader_1 {
	text-align: center;
}
``` 


